---
title: "Homework 1: Linear Algebra Technique"
author: "Kun-Hong Liao"
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: show
execute:
  warning: false
  message: false
---

## Setup and Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(Rcpp)
library(scatterplot3d)

sourceCpp("../01_linearAlgebra/matrixDecomposition/src/luDecomposition.cpp")
sourceCpp("../01_linearAlgebra/matrixDecomposition/src/qrDecomposition.cpp")
sourceCpp("../01_linearAlgebra/eigenvalueProblem/powerIteration.cpp")
sourceCpp("../01_linearAlgebra/eigenvalueProblem/qrEigenSolver.cpp")
```

# Data Loading and Preprocessing

```{r}
lines <- readLines("../data/CAmaxTemp.txt", warn = FALSE)
CA <- matrix(NA_real_, nrow = 12, ncol = 12)
for (i in seq_along(lines))
{
  parts <- unlist(strsplit(lines[i], "\\s+"))
  parts <- parts[parts != ""]
  
  n <- length(parts)
  
  CA[i, ] <- as.numeric(parts[(n - 12): (n - 1)]) 
}
CA
```

# 1. Monthly Record-Breaking Temperature in California I: Matrix Calculation

## (a) LU Factorization

Run the custom-implemented LU decomposition to obtain the lower triangular matrix $L$ and upper triangular matrix $U$. 

```{r}
#| label: q1-(a)
options(width = 200)
resLU <- LUDecompose(CA)

# The upper triangular matrix (4 digits)
L <- resLU$L
round(L, 4)

# The lower triangular matrix (4 digits)
U <- resLU$U
round(U, 4)
```

## (b) QR Factorization

Run the custom-implemented QR decomposition to obtain the orthogonal matrix $Q$ and upper triangular matrix $R$.

```{r}
#| label: q1-(b)
options(width = 200)
resQR <- QRDecompose(CA)

# The orthogonal matrix (4 digits)
Q <- resQR$Q
round(Q, 4)

# The upper triangular matrix (4 digits)
R <- resQR$R
round(R, 4)

# The inverse of CA matrix via QR decomposition
invQR <- QRInvert(resQR)
round(invQR, 4)
```

## (c) Power Iteration

Using the custom-implemented power iteration method to find the largest eigenvalue-eigenvector pair of CA matrix.

```{r}
#| label: q1-(c)
options(width = 200)

resPower <- PowerIteration(CA)

# The largest eigenvalue
resPower$eigenvalue

# The largest eigenvector
round(resPower$eigenvector, 4)
```

## (d) QR Algorithm

```{r}
#| label: q1-(d)

options(width = 200)

resQREigen <- QREigenSolver(CA, maxIter = 500)
```

# 2. Monthly Record-Breaking Temperature in California II: PCA and ICA

# 3. Monthly Record-Breaking Temperature in California III: ICA

# 4. Determinant and Parallogram

# Appendix: C++ Source Code

The following sections present the core C++ source codes used in this report. To maintain a concise layout, 
the code blocks are folded by default. Please click to expand and review the underlying computational logic. 

### 1. `luDecomposition.cpp`

```{Rcpp}
//| label: cpp-lu
//| eval: false
//| code-fold: true
//| code-summary: "Click to expand LU Decomposition source code"

#include <Rcpp.h>
#include <cmath>
using namespace Rcpp; 

/**
 * LU Decomposition with Partial Pivoting (Doolittle's Method)
 * * This function decomposes a square matrix A into P * A = L * U, where:
 * - L is a unit lower triangular matrix (diagonal elements are 1)
 * - U is an upper triangular matrix
 * - P is a permutation matrix recording row swaps
 *
 * @param mat A square NumericMatrix (NxN)
 * @return A List containing:
 * - "L": The lower triangular matrix (NxN)
 * - "U": The upper triangular matrix (NxN)
 * - "P": The permutation matrix (NxN)
 * - "swaps": Integer count of row swaps performance 
 */

// [[Rcpp::export]]
List LUDecompose(NumericMatrix mat){
  int n = mat.nrow();
  if(mat.ncol() != n) stop("Matrix must be square"); 
  
  NumericMatrix A = clone(mat); 
  
  //Initial permutation matrix
  NumericMatrix P_mat(n, n); 
  for(int i = 0; i < n; i++){
    P_mat(i, i)=1;
  }
  int swaps = 0; 
  for(int i = 0; i < n; i++){
    
    // pivoting
    double maxVal = 0.0;
    int imax = i; 
    
    for(int k = i; k < n; k++){
      if(std::abs(A(k, i)) > maxVal){
        maxVal = std::abs(A(k, i)); 
        imax = k; 
      }
    }
    
    if(maxVal < 1e-12) stop("Matrix is singular"); 
    
    // row swapping
    if(imax != i){
      swaps++; 
      for(int col = 0; col < n; col++){
        double temp = A(i, col); 
        A(i, col) = A(imax, col); 
        A(imax, col) = temp; 
      }
      
      for(int col = 0; col < n; col++){
        double temp = P_mat(i, col); 
        P_mat(i, col) = P_mat(imax, col);
        P_mat(imax, col) = temp; 
      }
    }
    
    // Doolittle Method
    for(int j = i + 1; j < n; j++){
      // L matrix
      A(j, i) /= A(i, i); 
      
      for(int k = i + 1; k < n; k++){
        // U matrix
        A(j, k) -= A(j, i)*A(i, k);
      }
    }
  }
  
  NumericMatrix L(n, n); 
  NumericMatrix U(n, n); 
  
  for(int i = 0; i < n; i++){
    for(int j = 0; j < n; j++){
      if(i > j){
        // lower triangular
        L(i, j) = A(i, j); 
        U(i, j) = 0; 
      } else if(i == j) {
        // diagonal
        L(i, j) = 1; 
        U(i, j) = A(i, j); 
      } else {
        // upper triangular
        L(i, j) = 0; 
        U(i, j) = A(i, j); 
      }
    }
  }
  
  return List::create(
    Named("L") = L, 
    Named("U") = U, 
    Named("P") = P_mat, 
    Named("swaps") = swaps
  ); 
}
```

### 2. `QRDecompose.cpp`

```{Rcpp}
//| label: cpp-qr
//| eval: false
//| code-fold: true
//| code-summary: "Click to expand QR Decomposition and inverse source code"

#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

/**
 * QR decomposition using Modified Gram-Schmidt
 */
// [[Rcpp::export]]
List QRDecompose(NumericMatrix mat){
  int m = mat.nrow(); // rows
  int n = mat.ncol(); // cols
  
  NumericMatrix Q = clone(mat); 
  NumericMatrix R(n, n); 
  
  for(int k = 0; k < n; k++){
    double norm_sq = 0.0; 
    for(int i = 0; i < m; i++){
      norm_sq += Q(i, k)*Q(i, k); 
    }
    R(k, k) = std::sqrt(norm_sq); 
    
    for(int i = 0; i < m; i++){
      Q(i, k) /= R(k, k); 
    }
    
    for(int j = k + 1; j < n; j++){
      double dot = 0.0; 
      for(int i = 0; i < m; i++){
        dot += Q(i, k)*Q(i, j); 
      }
      R(k, j) = dot; 
      
      for(int i = 0; i < m; i++){
        Q(i, j) -= R(k, j) * Q(i, k);
      }
    }
  }
  
  return List::create(
    Named("Q") = Q, 
    Named("R") = R
  ); 
}

/**
 * Calculate the Inverse Matrix using QR Decomposition
 * 
 */
// [[Rcpp::export]]
NumericMatrix QRInvert(List qr_result){
  NumericMatrix Q = qr_result["Q"]; 
  NumericMatrix R = qr_result["R"];
  
  int m = Q.nrow(); //row
  int n = Q.ncol();
  if(m != n) stop("Matrix must be square. ");
  
  NumericMatrix IA(n, n); 
  
  for(int j = 0; j < n; j++){
    NumericVector y(n);
    for(int i = 0; i < n; i++){
      y[i] = Q(j, i);
    }
    
    for(int i = n - 1; i >= 0; i--){
      double sum = 0.0; 
      for(int k = i + 1; k < n; k++){
        sum += R(i, k)*IA(k, j); 
      }
      IA(i, j) = (y[i] - sum) / R(i, i); 
    }
  }
  
  return IA; 
}
```

### 3. `powerIteration.cpp`

```{Rcpp}
#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

//[[Rcpp::export]]
List PowerIteration (NumericMatrix A, int maxIter = 1000, double tol = 1e-12)
{
  int n = A.nrow();
  if (n != A.ncol()) stop("Matrix must be square");
  
  NumericVector v(n); 
  
  // Initialization
  for (int i = 0; i < n; i++)
  {
    v[i] = R::rnorm(0.0, 1.0); 
  }
  
  // L2 normalization
  double norm = 0.0; 
  for(int i = 0; i < n; i++) norm += v[i] * v[i];
  norm = std::sqrt(norm);
  for(int i = 0; i < n; i++) v[i] /= norm;
  
  double lambdaOld = 0.0; 
  double lambdaNew = 0.0; 
  
  // Main Loop
  for (int k = 0; k < maxIter; k++)
  {
    NumericVector w(n); 
    
    // matrix multiplier
    for (int i = 0; i < n; i++)
    {
      double sum = 0.0; 
      for (int j = 0; j < n; j++)
      {
        sum += A(i, j) * v[j]; 
      }
      w[i] = sum; 
    }
    
    // Rayleigh Quotient
    lambdaNew = 0.0; 
    for (int i = 0; i < n; i++)
    {
      lambdaNew += v[i] * w[i]; 
    }
    
    if (std::abs(lambdaNew - lambdaOld) < tol) 
    {
      break;
    }
    lambdaOld = lambdaNew;
    
    norm = 0.0;
    for(int i = 0; i < n; i++) norm += w[i] * w[i];
    norm = std::sqrt(norm);
    for(int i = 0; i < n; i++) v[i] = w[i] / norm;
  }
  
  return List::create(
    Named("eigenvalue") = lambdaNew,
    Named("eigenvector") = v
  );
}
```

