---
title: "Homework 1: Linear Algebra Technique"
author: "Kun-Hong Liao"
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: show
execute:
  warning: false
  message: false
---

## Setup and Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(Rcpp)
library(scatterplot3d)

sourceCpp("../01_linearAlgebra/matrixDecomposition/src/luDecomposition.cpp")
sourceCpp("../01_linearAlgebra/matrixDecomposition/src/qrDecomposition.cpp")
sourceCpp("../01_linearAlgebra/eigenvalueProblem/powerIteration.cpp")
```

# Data Loading and Preprocessing

```{r}
lines <- readLines("../data/CAmaxTemp.txt", warn = FALSE)
CA <- matrix(NA_real_, nrow = 12, ncol = 12)
for (i in seq_along(lines))
{
  parts <- unlist(strsplit(lines[i], "\\s+"))
  parts <- parts[parts != ""]
  
  n <- length(parts)
  
  CA[i, ] <- as.numeric(parts[(n - 12): (n - 1)]) 
}
CA
```

# 1. Monthly Record-Breaking Temperature in California I: Matrix Calculation

## (a) LU Factorization

Run the custom-implemented LU decomposition to obtain the lower triangular matrix $L$ and upper triangular matrix $U$. 

```{r}
#| label: q1-(a)
options(width = 200)
resLU <- LUDecompose(CA)

# The upper triangular matrix (4 digits)
L <- resLU$L
round(L, 4)

# The lower triangular matrix (4 digits)
U <- resLU$U
round(U, 4)
```

## (b) QR Factorization

Run the custom-implemented QR decomposition to obtain the orthogonal matrix $Q$ and upper triangular matrix $R$.

```{r}
#| label: q1-(b)
options(width = 200)
resQR <- QRDecompose(CA)

# The orthogonal matrix (4 digits)
Q <- resQR$Q
round(Q, 4)

# The upper triangular matrix (4 digits)
R <- resQR$R
round(R, 4)

# The inverse of CA matrix via QR decomposition
invQR <- QRInvert(resQR)
round(invQR, 4)
```

## (c) Power Iteration

Using the custom-implemented power iteration method to find the largest eigenvalue-eigenvector pair of CA matrix.

```{r}
#| label: q1-(c)
options(width = 200)

resPower <- PowerIteration(CA)

# The largest eigenvalue
resPower$eigenvalue

# The largest eigenvector
round(resPower$eigenvector, 4)
```

## (d) QR Algorithm


